---
title: "Reproducible Research: Peer Assessment 1"
author: "De-Mei Meng"
date: "June 16, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
Download and unzip the Activity monitoring data file into the working directory

1.Load the data
```{r read data}
activity <- read.csv("activity.csv")
```
2.Process/transform the data
```{r process data}
activity$date<- as.Date(activity$date, format="%Y-%m-%d")
```
## What is mean total number of steps taken per day?
Ignore the missing values in the dataset.
```{r total steps per day}
steps_per_day <- aggregate(activity$steps ~ activity$date, FUN = "sum", rm.na = TRUE)
colnames(steps_per_day)<- c("Date", "Steps")
```
1.Make a histogram of the total number of steps taken each day
```{r histogram total steps per day}
hist(steps_per_day$Steps, xlab = "Steps", main = "Total Number of Steps Per Day")
```

2.Calculate and report the mean and median total number of steps taken per day
```{r mean}
mean(steps_per_day$Steps)
```

```{r median}
median(steps_per_day$Steps)
```
## What is the average daily activity pattern?
```{r average steps per interval}
average_steps <- aggregate(activity$steps ~ activity$interval, FUN = "mean", rm.na = TRUE)
colnames(average_steps)<- c("Interval", "Steps")
```
1.Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r plot for average steps per interval}
plot(average_steps, type = "l", main = "Average Steps in 5-Minute Interval")
```

2.Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r interval with maximum average steps}
average_steps[which.max(average_steps$Steps), ]$Interval
```
## Imputing missing values
The presence of missing days may introduce bias into some calculations or summaries of the data.
 
1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r count missing value rows}
nrow(activity[is.na(activity$steps),])
```
2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
```{r }
# Replace NA with the mean for that 5-minute interval previously calculated
```
3.Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r fill missing value}
activity_imputed <- transform(activity, steps = ifelse(is.na(activity$steps), yes = average_steps$Steps, no = activity$steps))
```

4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 
```{r total steps per day with filled value}
steps_per_day_imputed <- aggregate(activity_imputed$steps ~ activity_imputed$date, FUN = "sum")
colnames(steps_per_day_imputed)<- c("Date", "Steps")
hist(steps_per_day_imputed$Steps, xlab = "Steps", main = "Total Number of Steps Per Day", col="Navy")
hist(steps_per_day$Steps, xlab = "Steps", main = "Total Number of Steps Per Day", col="Light Blue", add = T)
legend("topright", c("Imputed Data", "Non-NA Data"), fill=c("Navy", "Light Blue") )
```
```{r mean with filled data}
mean(steps_per_day_imputed$Steps)
```

```{r median with filled data}
median(steps_per_day_imputed$Steps)
```
Do these values differ from the estimates from the first part of the assignment? 

Yes.

What is the impact of imputing missing data on the estimates of the total daily number of steps?

It increases the frequency near the mean/median value.

## Are there differences in activity patterns between weekdays and weekends?
Use the dataset with the filled-in missing values for this part.

1.Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r }
## Create new category based on the days of the week
activity_imputed$weekday <- weekdays(as.Date(activity_imputed$date))
activity_imputed$category <- ifelse(activity_imputed$weekday %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
```
2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data:
```{r }
library(ggplot2)
average_steps_imputed <- aggregate(steps~interval + category, activity_imputed, mean)
plot<- ggplot(average_steps_imputed, aes(x = interval , y = steps, color = category)) +
       geom_line() +
       labs(title = "Average Steps in 5-Minute Interval by Day Type", x = "Interval", y = "Average number of steps") +
       facet_wrap(~category, ncol = 1, nrow=2)
print(plot)
```

The plots shows higher average steps in the afternoon during weekends, compared to those during weekdays.


